---
title: '6\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: "Michael Sioson; Quantitative Biodiversity, East Carolina University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the  worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, it is *imperative* that you **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your '8.BetaDiversity' folder.
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**8.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**8.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 8^th^, 2023 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/6.BetaDiversity*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list = ls())
print(getwd())
setwd("~/Spring2024/BIOL6414_Quantitative_Biodiversity")
library(vegan)



```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting
require("ade4")
data("doubs")


```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}
quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))
fish <- doubs$fish
fish <- fish[-8, ]
adonis2(fish ~ quality, method = "bray", permutations = 999)
install.packages("indicspecies")
require("indicspecies")
indval <- multipatt(fish, cluster = quality, func = "IndVal.g", control = how (nperm = 999))
summary(indval)
fish.rel <- decostand(fish, method = "total")
phi <- multipatt(fish.rel, cluster = quality, func = "r.g", control = how (nperm = 999))
summary(phi)



```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: Habitat quality explains 44% of variation in fish species composition and is unlikely to be due to chance (p = 0.001). Alburnus, Gobio, Rutilus, and Lepomis were the strongest indicators of low or medium quality, while Salmo was the strongest indicator of high quality according to IndVal. According to phi, Phoxinus and Salmo were the strongest indicators of high quality and Alburnus was the strongest indicator of low quality. These analyses are mostly consistent, with some difference in species. This agrees with the other visualizations (worksheet 1), as sites with Alburnus and Salmo were significantly different.


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
fish.dist <- vegdist(doubs$fish[-8, ], method = "bray")
env.dist <- vegdist(scale(doubs$env[-8, ]), method = "euclid")
mantel(fish.dist, env.dist)



```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***: Stream environmental conditions explain 60% of variation in fish diversity and is unlikely to be due to chance (p = 0.001). We can reject the null hypothesis that stream quality does not influence fish communities.

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}
env.chem <- as.matrix(doubs$env[-8, 5:11])
env.chem
fish.db <- vegdist(fish, method = "bray", upper = TRUE, diag = TRUE)
doubs.dbrda <- dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)
psych::corr.test(env.chem)
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))
ordiplot(doubs.dbrda.mod0)
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)
doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[,c(4,6,7)], perm = 999)
dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] /
sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] /
sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
par(mar = c(5, 5, 4, 4) + 0.1)
plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1),
ylim = c(-1.1, 2.7), xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)",
sep = ""), ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
pch = 16, cex = 2.0, type = "n", cex.lab = 1.5,
cex.axis = 1.2, axes = FALSE)
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
points(scores(doubs.dbrda, display = "wa"),
pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"),
labels = row.names(scores(doubs.dbrda, display = "wa")))
vectors <- scores(doubs.dbrda, display = "bp")
arrows(0, 0, vectors[,1], vectors[, 2],
lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[, 2], pos = 3,
labels = row.names(vectors))
axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))

```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***: Dissolved oxygen concentration, dissolved nitrate concentration, and biological oxygen demand.

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova
env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]
env.mod
rs <- rowSums(fish)/sum(fish)
doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)
doubs.pcnmw$values > 0
doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)
plot(step.pcnm)
step.pcnm$anova
space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 +
PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]
space.mod
doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))
permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)
doubs.varpart <- varpart(fish.db, env.mod, space.mod)
doubs.varpart
par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -3)
```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: 37% of variation of fish diversity is explained by both environmental factors and location, with 13% and 24% being explained by environmental factors and location, respectively. Variation that could be explained by neither was 27%.

## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

